---
title: "TrackSOM-workflow"
output: rmarkdown::html_vignette
author: Givanna Putri
vignette: >
  %\VignetteIndexEntry{TrackSOM-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this tutorial, we outline a step by step instruction on how to use TrackSOM to cluster and track time-series data.
For simplicity, we will use the synthetic dataset outlined in our manuscript uploaded to [bioArxiv](https://www.biorxiv.org/content/10.1101/2021.06.08.447468v1).
The dataset files are provided in the `inst` directory: [link](https://github.com/ghar1821/TrackSOM/tree/master/inst/extdata).

# Importing TrackSOM package

If you have not installed the TrackSOM package, please use `devtools` to install the package from the following [TrackSOM github repo](https://github.com/ghar1821/TrackSOM).
For devtools, the repo parameter will be: `ghar1821/TrackSOM`.

The following code shall import the TrackSOM package:
```{r message=FALSE, warning=FALSE}
library(TrackSOM)
```

# Importing dataset

TrackSOM supports dataset stored as either CSV or FCS files or as `data.table` object (enhanced version of R's native data.frame. See [data.table vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) for more details).
The following sections shall show you how to read in those files and pass them to the TrackSOM function.

## Dataset as CSV files

Here, we assume that each file contains data belonging to one time-point. 
Hence you should have more than 1 CSV files.
If this is not the case, please reformat your data files.

To import the dataset stored as CSV files, TrackSOM needs to know where the files are stored.
These files' location must be stored within a vector which get passed on to the TrackSOM function.

*Important:* the vector must be organised such as the first element is the data for the very first time-point, the 2nd element for the 2nd time-point, and so on.

First, we start with specifying the CSV files are.
In this example, the dataset files are already stored within the package, so all you need to do is load it up:
```{r}
data.files.fullpath <- c(
  system.file("extdata", "synthetic_d0.csv", package = "TrackSOM"),
  system.file("extdata", "synthetic_d1.csv", package = "TrackSOM"),
  system.file("extdata", "synthetic_d2.csv", package = "TrackSOM"),
  system.file("extdata", "synthetic_d3.csv", package = "TrackSOM"),
  system.file("extdata", "synthetic_d4.csv", package = "TrackSOM")
)
```

*For your dataset, please replace the content of the vector with the **absolute path** for the dataset files!*

Let's inspect the content of `data.files.fullpath`:

```{r}
print(data.files.fullpath)
```

You can see it contains a list of *absolute path* of multiple CSV files, each belonging to a time-point.

## Dataset as FCS files

The procedure is very similar to reading CSV files above, except we're going to be storing FCS files' path rather than CSV.

Again, here, we assume that each file contains data belonging to one time-point. 
Hence you should have more than 1 FCS files.
If this is not the case, please reformat your data files.

To import the dataset stored as FCS files, TrackSOM needs to know where the files are stored.
These files' location must be stored within a vector which get passed on to the TrackSOM function.
*Important:* the vector must be organised such as the first element is the data for the very first time-point, the 2nd element for the 2nd time-point, and so on.

First, we start with specifying the FCS files are.
In this example, the dataset files are already stored within the package, so all you need to do is load it up:
```{r}
data.files.fullpath.fcs <- c(
  system.file("extdata", "synthetic_d0.fcs", package = "TrackSOM"),
  system.file("extdata", "synthetic_d1.fcs", package = "TrackSOM"),
  system.file("extdata", "synthetic_d2.fcs", package = "TrackSOM"),
  system.file("extdata", "synthetic_d3.fcs", package = "TrackSOM"),
  system.file("extdata", "synthetic_d4.fcs", package = "TrackSOM")
)
```

*For your dataset, please replace the content of the vector with the **absolute path** for the dataset files!*

Let's inspect the content of `data.files.fullpath.fcs`:

```{r}
print(data.files.fullpath.fcs)
```

You can see it contains a list of *absolute path* of multiple FCS files, each belonging to a time-point.

## Dataset as `data.table` object

Sometimes, it is convenient to have the dataset stored as the `data.table` object files, e.g. when you need to run some code to preprocess your data using R!
As an example, supposed the synthetic dataset CSV files were already read in as `data.table` object prior to running TrackSOM (say you did some preliminary clean up or filtering).
What you need to do is organise them in a list such that each element is a `data.table` object for the dataset in a time-point.
*Important:* the list must be organised such as the first element is the data for the very first time-point, the 2nd element for the 2nd time-point, and so on.

```{r}

library(data.table)
data.files <- c(
  system.file("extdata", "synthetic_d0.csv", package = "TrackSOM"),
  system.file("extdata", "synthetic_d1.csv", package = "TrackSOM"),
  system.file("extdata", "synthetic_d2.csv", package = "TrackSOM"),
  system.file("extdata", "synthetic_d3.csv", package = "TrackSOM"),
  system.file("extdata", "synthetic_d4.csv", package = "TrackSOM")
)
dat <- lapply(data.files, function(f) fread(f))
```

Let's do a quick preview of the data:
```{r}
dat
```

As you can see, there are 5 elements in the list, each containing a dataset belonging to a time-point.

# Running TrackSOM

Depending on how your dataset is stored, the parameter `inputFiles` is either a vector of absolute path of your CSV or FCS files or a list of `data.table` object.
Additionally, you need to specify the type as the parameter `dataFileType`. 
It can be either `.csv`, `.fcs`, or `data.frame` depending on how your dataset is stored.

## Other parameters

***Note:*** examples here assume datasets are stored as CSV files.

The TrackSOM function have various parameters.
Of them, only `inputFiles` and `colsToUse` have no default values. 
The `inputFiles` parameter has been explained in the previous section.

The `colsToUse` parameter specify the columns in your dataset to be used for clustering and tracking.
This must be a vector.
For the synthetic dataset, the columns are denoted as `x`, `y`, and `z`. 
For cytometry data, this should be a vector of markers.

The TrackSOM functions have the following parameters which are pre-filled with default values:

* tracking: TRUE (default) or FALSE, whether to perform tracking.
* noMerge: TRUE or FALSE (default), whether to allow merging of meta-clusters (FALSE) or not (TRUE). If noMerge is set to TRUE, meta-clusters are only allowed to split, but not merge.
* maxMeta: NULL (default) or round number. Maximum number of meta clusters for all time point. Only used for Autonomous Adaptive operation.
* nClus: NULL (default) or numeric or a vector of numbers.  If single number, TrackSOM will run Prescribed Invariant operation. Otherwise, it will run Prescribed Variant operation.
* seed: 42 (default). Random seed number.
* xdim: 10 (default). SOM grid size.
* ydim: 10 (default). SOM grid size.

The function also accept parameters that are built into [FlowSOM](https://bioconductor.org/packages/release/bioc/html/FlowSOM.html) `ReadInput`, `BuildSOM` and `BuildMST` functions.
See FlowSOM's vignette for specific parameter information.

Let's run TrackSOM with the following settings:

* No merging of meta-clusters allowed (`noMerge = TRUE`)
* Enable tracking (`tracking = TRUE`)
* Prescribed Variant producing 3, 3, 9, 7, 15 meta-clusters for each time-point.

The remaining parameters will be set to the default values.
```{r}
tracksom.result <- TrackSOM(inputFiles = data.files.fullpath,
                            colsToUse = c('x', 'y', 'z'),
                            tracking = TRUE,
                            noMerge = TRUE,
                            nClus = c(3,3,9,7,15),
                            dataFileType = ".csv"
)
```

# Extracting meta-clusters ID and SOM nodes

TrackSOM result is stored in an object. 
To facilitate the extraction of meta-clusters ID and SOM nodes for each cell, we provide the following functions:

* `ConcatenateClusteringDetails`: assuming your dataset is stored as a `data.table` object, the function attaches the meta-cluster ID and SOM nodes as separate columns.
* `ExportClusteringDetailsOnly`: this function simply extract the meta-cluster ID and SOM nodes for each cell as a `data.table` object. The cells are ordered based on the ordering of your dataset. 

To use `ConcatenateClusteringDetails`, you need to first read in all your datasets as one giant `data.table`.
```{r}
library(data.table)
InputDirectory <- "~/Documents/GitHub/TrackSOM/inst/extdata/"
data.files <- list.files(InputDirectory, ".csv")
dat <- lapply(data.files, function(f) fread(paste0(InputDirectory, f)))
dat <- rbindlist(dat)
```

Double check the data is read in properly:
```{r}
head(dat)
```
```{r}
tail(dat)
```

To run `ConcatenateClusteringDetails`, you need to pass the following parameters:

* `timepoint.col`: which column in your data define the time-point.
* `timepoints`: what are the time-points (in order). This must be a vector.

Now let's attach the meta-cluster ID and the SOM nodes:
```{r}
dat.clust <- ConcatenateClusteringDetails(
    tracksom.result = tracksom.result,
    dat = dat,
    timepoint.col = "timepoint",
    timepoints = c('Mock', 'SYN-1', 'SYN-2', 'SYN-3', 'SYN-4')
    )
```

Inspect the content:
```{r}
head(dat.clust)
```
The function attaches extra 3 columns:

* `TrackSOM_cluster`: The SOM node of each cell.
* `TrackSOM_metacluster`: The meta-cluster assignment produced by FlowSOM's meta-clustering.
* `TrackSOM_metacluster_lineage_tracking`: The tracking of meta-clusters' evolution produced by TrackSOM. **This gives you the changes undergone by the meta-clusters over time**.

The function gives back a `data.table` object which you can export as CSV file using `data.table`'s `fwrite` function.

# Visualisations

TrackSOM offers 2 visualisation mediums:

1. Network plot.
2. Timeseries heatmap.

## Network plot

To draw network plots, we need to call the `DrawNetworkPlot` function.
Using the clustered and tracked data from previous sections, the following is an example on how to use the function:
```{r}
DrawNetworkPlot(dat = dat.clust,
                timepoint.col = "timepoint",
                timepoints = c('Mock', 'SYN-1', 'SYN-2', 'SYN-3', 'SYN-4'),
                cluster.col = 'TrackSOM_metacluster_lineage_tracking',
                marker.cols = c('x', 'y', 'z'))
```

The `marker.cols` will determine the *markers* which mean/median expression will be drawn on the network plots.
There is no need to specify all the markers in the dataset, just the ones that you want the network plots to be coloured on.

The function won't preview any plots, but it will instead store the plots as image files and some extra information (median/mean expression of markers) as CSV files:
```{r}
list.files()
```

In this example, the plots are saved as PDF files.
This can be changed, e.g. to save the plots as PNG files, by specifying the desired file format as the `file.format` parameter.

## Timeseries heatmap

To draw a timeseries heatmap, we need to call the `DrawTimeseriesHeatmap` function.
Using the clustered and tracked data from previous sections, the following is an example on how to use the function:
```{r}
DrawTimeseriesHeatmap(dat = dat.clust,
                timepoint.col = "timepoint",
                timepoints = c('Mock', 'SYN-1', 'SYN-2', 'SYN-3', 'SYN-4'),
                cluster.col = 'TrackSOM_metacluster_lineage_tracking',
                marker.cols = c('x', 'y', 'z'))
```

The `marker.cols` will determine the *markers* which mean/median expression will be drawn on the network plots.
There is no need to specify all the markers in the dataset, just the ones that you want the network plots to be coloured on.

The function won't preview any plots, but it will instead store the plots as image files and some extra information (median/mean expression of markers) as CSV files:
```{r}
list.files()
```

In this example, the plots are saved as PDF files.
This can be changed, e.g. to save the plots as PNG files, by specifying the desired file format as the `file.format` parameter.

# Concluding remarks

That is pretty much it folks!
We're actively updating TrackSOM and welcome feedbacks! 

Thank you for your interest!
